How is experience transformed into memory?

### 2018-08-23 paper/main.tex
supp mat formatting
moved refs above methods
JRM abstract edits

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
% \usepackage{amsmath}
% \newcommand{\rpm}{\raisebox{.2ex}{$\scriptstyle\pm$}}
% \DeclareMathOperator*{\argmax}{arg\,max}
\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Supplementary Figure},fileext=lof]{suppfigure}

\title{Capturing the shape of our experiences and how we remember them}
\author{Andrew C. Heusser \& Jeremy R. Manning}
% \affiliation{Dartmouth College}

@@

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner. This temporal structure gives each of our experiences (and thus our memories for those experiences) a unique shape.  However, temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the temporal structure (i.e. the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. Using a novel approach based on topic modelling, we found that despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

\section{Introduction}

@@

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  For instance, naturalistic experiences are autocorrelated in space and time on short timescales. In other words, the contents of an experience are often similar from moment to moment and change gradually. As an example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall the isolated elements of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e. correlation) between their topic vectors. In other words, the video event that is most similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}f). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g for example, Supp. Fig~\ref{fig:matchmats} for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e. a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow.}
\label{fig:trajectory}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event (per subject). Error bars represent 95\% confidence intervals. B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g. the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

\section{Discussion}
\label{sec:discussion}
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

@@

In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g. percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. We view ``true learning'' as understanding the key concepts (i.e. understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

@@

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g. one that describes the video in the same way as participant's recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g. participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to details (while maintaining flexibility).

@@

\section{Methods}
\label{sec:methods}

@@

\subsubsection{Fitting the topic model to the video text and recall transcripts}
A topic model was used to estimate the most likely mixture of topics for a given sample of text. First, the video was manually segmented into 1000 scenes, and a collection of descriptive features was manually transcribed. For each scene, we considered the following features: scene details (i.e. a sentence or two describing what happened in that scene), space (indoor or outdoor), name of all the characters in the scene, name of the character in focus, name of the character speaking, location, camera angle, music presence, and words on the screen. We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene. We then transformed the text descriptions into overlapping windows of 50 scene segments. For example, the first text sample comprised the text from the first 50 segments, the second comprised the text from n+1:n+51, and so on. We trained our model using these overlapping text samples with scikit-learn's (version 0.19.1) `CountVectorizer' and `LatentDirichletAllocation' classes \citep{PedrEtal11} implemented using our high-dimensional visualization/analysis software, Hypertools \citep{HeusEtal17a}.  First, the text was transformed into a vector of word counts (after removing English stopwords). This gave a word count vector for each scene in the video.  Then, the word count vectors were used to fit a topic model (topics=100, method=`batch'). We transformed the text descriptions using the model resulting in a scenes (1000) by topics (100) matrix. The scene descriptions often spanned multiple timepoints (i.e. TRs). To account for this, we expanded the video model by copying the rows of the model for as many timepoints that the scene description spanned. After this expansion, the shape of the model was the length of the duration of the video (1976 TRs).

@@

\subsubsection{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g. hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

\subsubsection{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events, e.g. Supp. Fig.~\ref{fig:corrmats}). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events ~\citep{Rabi89, BaldEtal17}. To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17, Brainiak}.

Our metric for choosing the ``best fitting'' HMM was to choose the model with the $k$ value that maximized the ratio of the average `within-event' correlation values (i.e. the correlation values for blocks of consecutive timepoints the model identified as one event) and the average `across-event' correlation (i.e. the rest of the correlation values). Additionally, we included a penalty parameter that was proportional to the smoothing of the model that preferred models with smaller $k$ values. We chose $k$ values separately for the video model and for each recall model.  Then, using the best $k$ values, We fit a separate HMM for the video and each recall model. Finally, we averaged over timepoints identified to be in the same event resulting in a events by topics matrix for the video model and each of the recall models.

@@

\subsubsection{Vector field analysis}
To quantify the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a circle around the point (radius=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e. the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e. displayed a preferred direction across participants), we performed a Rayleigh test on the angles ($p<.001$, FDR-corrected at $p<.05$ using Benjamani-Hochberg procedure). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.

@@

\bibliography{memlab}

@@

\section{Naturalistic extensions of classic list-learning analyses}
Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. Here, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings. First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r(16)=.67, p=.003$). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments ~\citep{HowaKaha99}. Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}c). We did not observe the classic ``primacy'' and ``recency'' pattern which is prevalent in the literature ~\citep{Murd62a}. We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r(16)=.62, p=.007$). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r(16)=.55, p=.02$).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Supplementary Figure},fileext=lof]{suppfigure}

\title{How is experience transformed into memory?}
\author{Andrew C. Heusser and Jeremy R. Manning\\Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that, despite the rich complexities of real-world experiences, they unfold in a structured and often predictable manner. This structure gives each of our experiences (and thus our memories for those experiences) a unique `shape' with respect to how it unfolds in the relevant representational space.  However, temporal structure is typically absent (or severely impoverished) in traditional trial-based  experiments. As a consequence, our understanding of how the temporal structure (i.e., the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the video to participants' later recounting of the narrative. Using a novel approach that leverages topic modeling and Hidden Markov Models, we found that, despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by nearly every participant. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.
\end{abstract}

\section*{Introduction}

@@

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e., how the contents relate/change over time).  For instance, naturalistic experiences are autocorrelated in space and time on short timescales. In other words, the contents of an experience are often similar from moment to moment and change gradually. As an example, many of us spend our mornings/evenings at home and much of our days at work (e.g.\ an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall the isolated elements of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our experiences often change gradually, they can also change rapidly (e.g.\ walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}f). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g for example, Supp. Fig~\ref{fig:matchmats} for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e., the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow.}
\label{fig:trajectory}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event (per subject). Error bars represent 95\% confidence intervals. B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

\section*{Discussion}
\label{sec:discussion}
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

@@

In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. We view ``true learning'' as understanding the key concepts (i.e., understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

@@

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g.\ one that describes the video in the same way as participant's recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g.\ participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to details (while maintaining flexibility).

@@

\pagebreak
\bibliography{memlab}

\pagebreak
\section*{Methods}
\label{sec:methods}

@@

\subsubsection{Fitting the topic model to the video text and recall transcripts}
A topic model was used to estimate the most likely mixture of topics for a given sample of text. First, the video was manually segmented into 1000 scenes, and a collection of descriptive features was manually transcribed. For each scene, we considered the following features: scene details (i.e., a sentence or two describing what happened in that scene), space (indoor or outdoor), name of all the characters in the scene, name of the character in focus, name of the character speaking, location, camera angle, music presence, and words on the screen. We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene. We then transformed the text descriptions into overlapping windows of 50 scene segments. For example, the first text sample comprised the text from the first 50 segments, the second comprised the text from n+1:n+51, and so on. We trained our model using these overlapping text samples with scikit-learn's (version 0.19.1) `CountVectorizer' and `LatentDirichletAllocation' classes \citep{PedrEtal11} implemented using our high-dimensional visualization/analysis software, Hypertools \citep{HeusEtal17a}.  First, the text was transformed into a vector of word counts (after removing English stopwords). This gave a word count vector for each scene in the video.  Then, the word count vectors were used to fit a topic model (topics=100, method=`batch'). We transformed the text descriptions using the model resulting in a scenes (1000) by topics (100) matrix. The scene descriptions often spanned multiple timepoints (i.e., TRs). To account for this, we expanded the video model by copying the rows of the model for as many timepoints that the scene description spanned. After this expansion, the shape of the model was the length of the duration of the video (1976 TRs).

@@

\subsubsection{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g.\ hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

\subsubsection{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events, e.g.\ Supp. Fig.~\ref{fig:corrmats}). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events ~\citep{Rabi89, BaldEtal17}. To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17, Brainiak}.

Our metric for choosing the ``best fitting'' HMM was to choose the model with the $k$ value that maximized the ratio of the average `within-event' correlation values (i.e., the correlation values for blocks of consecutive timepoints the model identified as one event) and the average `across-event' correlation (i.e., the rest of the correlation values). Additionally, we included a penalty parameter that was proportional to the smoothing of the model that preferred models with smaller $k$ values. We chose $k$ values separately for the video model and for each recall model.  Then, using the best $k$ values, We fit a separate HMM for the video and each recall model. Finally, we averaged over timepoints identified to be in the same event resulting in a events by topics matrix for the video model and each of the recall models.

@@

\subsubsection{Vector field analysis}
To quantify the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a circle around the point (radius=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e., the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e., displayed a preferred direction across participants), we performed a Rayleigh test on the angles ($p<.001$, FDR-corrected at $p<.05$ using Benjamani-Hochberg procedure). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.

@@

@@

\section*{Naturalistic extensions of classic list-learning analyses}
Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. Here, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r(16)=.67, p=.003$). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments ~\citep{HowaKaha99}. Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}c). We did not observe the classic ``primacy'' and ``recency'' pattern which is prevalent in the literature ~\citep{Murd62a}. We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r(16)=.62, p=.007$). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r(16)=.55, p=.02$).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.
